---
title: "Houston Labour-Market Time Series"
subtitle: "A corrected quarterly forecasting comparison"
author: "Reway Du"
date: "Portfolio edition — August 2026"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: hide
    theme: flatly
    self_contained: true
    pandoc_args: ["--resource-path=.."]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.align = "center",
  out.width = "95%"
)
project_root <- if (file.exists("run.R")) {
  normalizePath(
    ".",
    winslash = "/",
    mustWork = TRUE
  )
} else if (file.exists(file.path("..", "run.R"))) {
  normalizePath(
    "..",
    winslash = "/",
    mustWork = TRUE
  )
} else {
  stop("Cannot locate the repository root containing run.R.")
}

project_file <- function(...) {
  file.path(project_root, ...)
}

metadata <- read.csv(
  project_file("results", "analysis_metadata.csv"),
  stringsAsFactors = FALSE
)

selection <- read.csv(
  project_file("results", "model_selection.csv"),
  stringsAsFactors = FALSE
)

diagnostics <- read.csv(
  project_file("results", "residual_diagnostics.csv"),
  stringsAsFactors = FALSE
)

stationarity <- read.csv(
  project_file("results", "stationarity_tests.csv"),
  stringsAsFactors = FALSE
)

metrics <- read.csv(
  project_file("results", "forecast_metrics.csv"),
  stringsAsFactors = FALSE
)

unemployment_candidates <- read.csv(
  project_file("results", "candidate_models_unemployment.csv"),
  stringsAsFactors = FALSE
)

wage_candidates <- read.csv(
  project_file("results", "candidate_models_wage.csv"),
  stringsAsFactors = FALSE
)

metadata_values <- setNames(metadata$value, metadata$field)
selection_values <- setNames(selection$selected_value, selection$component)

metric_value <- function(target_name, model_name, metric_name) {
  metrics$value[
    metrics$target == target_name &
      metrics$model == model_name &
      metrics$metric == metric_name
  ][1]
}

fmt_num <- function(x, digits = 2L) {
  format(round(as.numeric(x), digits), nsmall = digits, big.mark = ",")
}

unemployment_target <- "Quarterly unemployment rate"
wage_target <- "Average weekly wage (USD)"
```

# Executive summary

This project revisits a two-part university time-series assignment using a
clean, frequency-correct pipeline. It compares separate seasonal ARMA models
with a two-variable VAR and a seasonal-naive baseline for one-quarter-ahead
forecasts of Houston-area unemployment and average weekly wages.

The analysis uses `r metadata_values[["aligned_observations"]]` aligned quarters
from `r metadata_values[["aligned_start"]]` through
`r metadata_values[["aligned_end"]]`. Models are initially selected using data
through `r metadata_values[["initial_training_end"]]`; the final
`r metadata_values[["evaluation_quarters"]]` quarters are evaluated with an
expanding training window.

The univariate models produced the lowest RMSE for both series. Unemployment
RMSE was `r fmt_num(metric_value(unemployment_target, "Univariate SARMA", "RMSE"))`
percentage points, versus
`r fmt_num(metric_value(unemployment_target, "Seasonal naive", "RMSE"))` for the
seasonal-naive baseline and
`r fmt_num(metric_value(unemployment_target, "VAR", "RMSE"))` for the VAR. Wage
RMSE was $`r fmt_num(metric_value(wage_target, "Univariate SARMA", "RMSE"))`,
versus $`r fmt_num(metric_value(wage_target, "Seasonal naive", "RMSE"))` and
$`r fmt_num(metric_value(wage_target, "VAR", "RMSE"))`, respectively.

This is evidence about forecast accuracy under one historical evaluation—not a
causal conclusion and not proof that wages and unemployment are unrelated.

# Data and frequency alignment

The two public FRED series have different native frequencies and units:

- [`HOUS448URN`](https://fred.stlouisfed.org/series/HOUS448URN) is a monthly,
  not-seasonally-adjusted unemployment rate in percent.
- [`ENUC264240010`](https://fred.stlouisfed.org/series/ENUC264240010) is a
  quarterly, not-seasonally-adjusted average weekly wage in dollars per week.

For every retained quarter, the unemployment rate is the arithmetic mean of
all three monthly observations. The wage observation is then joined by the same
quarter date. No vector recycling, row-position matching, or unequal-length
`cbind` is used.

```{r aligned-series, echo=FALSE, fig.cap="Aligned quarterly unemployment and average weekly wage series."}
knitr::include_graphics(
  project_file("figures", "01_aligned_series.png"),
  rel_path = FALSE
)
```

The wage series is not a minimum-wage measure. It covers average weekly wages
paid by unemployment-insurance-covered employers. The two source titles also
use different historical Houston MSA names; this analysis does not assume that
geographic definitions remained perfectly unchanged.

# Transformations and stationarity diagnostics

The modeled annual transformations are

$$
\Delta_4 u_t = u_t-u_{t-4}, \qquad
g^{(4)}_t = \log(w_t)-\log(w_{t-4}),
$$

where four lags correspond to one year of quarterly data. A lag of 12 would be
three years and is not used as a quarterly seasonal difference.

```{r transformed-series, echo=FALSE, fig.cap="Year-over-year transformations used for modeling."}
knitr::include_graphics(
  project_file("figures", "02_stationary_transformations.png"),
  rel_path = FALSE
)
```

```{r stationarity-table, echo=FALSE}
stationarity_display <- stationarity
stationarity_display$ADF_p_value <- format.pval(
  stationarity_display$ADF_p_value,
  digits = 3,
  eps = 0.001
)
stationarity_display$KPSS_level_p_value <- format.pval(
  stationarity_display$KPSS_level_p_value,
  digits = 3,
  eps = 0.001
)
knitr::kable(
  stationarity_display,
  col.names = c("Series", "N", "ADF p-value", "KPSS-level p-value"),
  align = c("l", "r", "r", "r"),
  caption = "Complementary stationarity diagnostics on the frozen snapshot."
)
```

The ADF null is a unit root, while the KPSS-level null is level stationarity.
For both annual transformations, the ADF test rejects its null and the KPSS test
does not reject its null at 5%. These tests have limited power and are treated as
diagnostics, not proof that the data-generating process is fixed forever.

# Model selection

Univariate candidates span non-seasonal orders $p,q\in\{0,1,2,3\}$ and seasonal
orders $P,Q\in\{0,1\}$ with quarterly period four and at most four total AR/MA
terms. Selection uses only the initial training sample. A candidate must have a
12-lag Ljung–Box p-value of at least 0.05, after adjusting for fitted AR/MA
terms; among passing candidates, the lowest AICc is selected.

```{r selection-table, echo=FALSE}
knitr::kable(
  selection,
  col.names = c("Component", "Selected value"),
  align = c("l", "l"),
  caption = "Initial-training model choices."
)
```

```{r diagnostic-table, echo=FALSE}
diagnostic_display <- diagnostics
diagnostic_display$Ljung_Box_statistic <- round(
  diagnostic_display$Ljung_Box_statistic,
  2
)
diagnostic_display$Ljung_Box_p_value <- round(
  diagnostic_display$Ljung_Box_p_value,
  3
)
knitr::kable(
  diagnostic_display,
  col.names = c(
    "Series", "Selected model", "Lag", "Ljung–Box statistic",
    "p-value", "Conclusion"
  ),
  align = c("l", "l", "r", "r", "r", "l"),
  caption = "Residual-autocorrelation checks on the initial training fits."
)
```

A Ljung–Box p-value below 0.05 would indicate remaining residual
autocorrelation; it would not establish stationarity. The selected models pass
this initial check, although passing one diagnostic does not guarantee correct
specification.

# Out-of-sample design

For every quarter from `r metadata_values[["evaluation_start"]]` through
`r metadata_values[["evaluation_end"]]`, each model is refit using only data
available before that target quarter. All methods make the same one-step-ahead
point forecast:

1. the seasonal-naive forecast repeats the same quarter from the previous year;
2. each SARMA model forecasts its own annual transformation and reconstructs
   the level from the same quarter one year earlier; and
3. the VAR jointly forecasts the two annual transformations and reconstructs
   both levels in the same way.

This avoids comparing a rolling one-step forecast from one model with a fixed-
origin multi-step forecast from another.

```{r forecast-figure, echo=FALSE, fig.cap="Expanding-window, one-quarter-ahead forecasts."}
knitr::include_graphics(
  project_file("figures", "03_rolling_forecasts.png"),
  rel_path = FALSE
)
```

```{r metric-table, echo=FALSE}
metric_display <- metrics
metric_display$value <- round(metric_display$value, 3)
knitr::kable(
  metric_display,
  col.names = c("Target", "Model", "Metric", "Value"),
  align = c("l", "l", "l", "r"),
  caption = "Point-forecast error over the 29-quarter evaluation."
)
```

```{r rmse-figure, echo=FALSE, fig.cap="RMSE comparison using a common evaluation design."}
knitr::include_graphics(
  project_file("figures", "04_rmse_comparison.png"),
  rel_path = FALSE
)
```

The univariate models lead on both MAE and RMSE. The VAR does not improve these
point forecasts, and it is less accurate than the seasonal-naive baseline for
unemployment. The result is stated directly rather than selecting whichever
metric or label produces the preferred story.

# Limitations

- The evaluation has only 29 quarters and includes the unusually large 2020
  labour-market disruption; forecast rankings may differ in a calmer period.
- A fixed 80/20 chronological split is transparent but does not quantify
  uncertainty in model-ranking differences.
- ADF, KPSS, AICc, and Ljung–Box checks each answer narrow questions and cannot
  prove model truth.
- The selected wage model's initial Ljung–Box p-value is only about 0.052, very
  close to the 0.05 screen; its residual adequacy should therefore be described
  as borderline rather than conclusive.
- FRED/BLS observations can be revised after the frozen download.
- The series are not seasonally adjusted and may reflect changing metropolitan
  definitions and reporting practices.
- The VAR is a forecasting benchmark. No structural identification, exogenous
  controls, Granger-causality study, or causal interpretation is attempted.
- Only point forecasts are scored; production use would also require calibrated
  forecast intervals and monitoring for structural breaks.

# Conclusion

Correct frequency alignment changes the project from an invalid monthly/
quarterly comparison into a reproducible forecasting exercise. On this frozen
sample and evaluation window, separate seasonal ARMA models forecast both
targets more accurately than the fitted VAR and seasonal-naive alternatives.
The most defensible conclusion is therefore modest: a joint linear dynamic model
did not add point-forecast value here.

# Portfolio and authorship note

The earlier two-part submission was completed collaboratively by Reway Du and
three peers. It contains no task-level contribution table, so this rewrite does
not assign sole original authorship to Reway. Student numbers, assignment
prompts, teammate names, and the original AI-percentage table are omitted.


# Computing environment

```{r session-info}
cat(
  readLines(
    project_file("results", "session-info.txt")
  ),
  sep = "\n"
)
```
